### 2017-12-24 Writing/draft.tex
minor changes following martin's comments

The last two decades have witnessed a resurgence of interest in the use of morphological character data in phylogenetic studies.
This owes in large part to the use of fossils to untertake at least partial reconstructions of phylogenetic trees, especially where ancestral states reconstructions or absolute calibrations of divergence times are necessary. 
Morphological character data are often considered inferior to molecular sequence data, but are often the only source of data---especially for fossils.
While there is a general appreciation of the limits of morphological data, they are frequently dismissed without any empirical investigations into their statistical properties.
As morphological data are likely to continue to play an extensive role in phylogenetic analysis, it is essential to understand the circumstances under which morphological data might be expected to 'misbehave'.
This opens up possibilities for predicting problematic datasets and possibly proposing new confidence measures in phylogenetic datasets.
%The last decade has witnessed a resurgence of use of discrete morphological data in macroevolutionary studies.
%Traditionally confined to ``classic'' cladistic studies in palaeontology, these methods have recently been widely expanded and improved whether based on the models and methods for inferring phylogenies \citep[e.g.][]{heath2014fossilized,Wright01072016} or to combine them with molecular data \citep[e.g.][]{pyrondivergence2011,ronquista2012}.
%In parallel, much development has also been done on the pre-phylogenetic analysis \citep[e.g. data collection;][]{morphobank} as well as on post-phylogenetic ones \citep[e.g. morphological disparity analysis;][]{Close2015,Claddis}.
%Nonetheless, using discrete morphological data in macroevolutionary studies comes with several caveats based to the nature of data used \citep{Guillerme2016146,bapst2017combined}; the coding of this data \citep{Brazeau2011,simoes2017giant}; the topology inference method \citep{spencerefficacy2013,wrightbayesian2014,OReilly20160081,puttick2017uncertain,goloboff2017weighted} or the dating methods \citep{Arcila2015131,o2016tips}.

@@

    \item \textbf{Evolutionary dependence:} this is the result of sets of characters co-evolving: the character's correlation is thus due to their shared evolutionary history.
% MDB: Does the above really describe this correctly. This actually sounds a lot more like genuine synapomorphy than potentially problematic correlation.

    Many methods have been developed to study these correlations, especially since they can provide us with a lot of information on how specific groups acquired specific characteristics \citep{Lande1983,Maddison1990,Pagel1994,Pagel2006,Grabowski2016}.
    For example, the apparition of the jaw as a character in gnatosthomes correlates with a whole suite of characters present in modern gnatosthomes (e.g. beaks, tribosphenic molars, jaw protrusion, etc.). %TG: this links to inapplicable data!
%MDB: The above is really coding dependence that you describe below, in a way. It's a character hierarchy. What we would like to discuss here is probably something more like limb reduction and axial elongation, for instance. These two things seem to co-vary extensively in tetrapods, with snake-like body forms evolving multiple times in numerous tetrapod lineages.

@@

\noindent Where $n$ is the number of taxa with comparable data and $x_i$ and $y_i$ are each characters states for the characters $x$ and $y$ and the taxa $i$.
$CD$ is a continuous distance metric bounded between $0$ and $1$ (see the mathematical demonstration in the Supplementary Material 1).
Since we are considering differences as being only Fitch-like (i.e. non-weighted and non-ordinate), we calculated the difference between character states in a qualitative way.
Two same character states tokens have a difference of zero and two different ones have a difference of one (e.g. $0 - 0 = 0$ or $1 - 8 = 1$).
Additionally, we only consider differences for taxa with shared information \citep[i.e. a Gower distance;][]{GowerDist}.
We standardised each character by arbitrarily modifying their character state tokens by order of appearance.
In other words, we replaced all the occurrences of the first token to be $1$, the second to be $2$, etc.
This way, a character \texttt{A = \{2,2,3,0,0,3\}} would be standardised as \texttt{A' = \{1,1,2,3,3,2\}} \citep[following the \textit{xyz} notation in][p.13]{felsenstein2004inferring}.
Note that in terms of phylogenetic signal, both \texttt{A} and \texttt{A'} are exactly identical (forming three distinct groups).

@@

\subsection{Simulating discrete morphological matrices}
To simulate the matrices we applied a protocol very similar to \cite{Guillerme2016146}.
First, we generate random birth-death trees with the birth ($\lambda$) and death ($\mu$) parameters sampled from a uniform $(0,1)$ distribution maintaining $\lambda$ $>$ $\mu$ using the \texttt{diversitree} \texttt{R} package \citep[v0.9-8;][]{fitzjohndiversitree2012} and saving the tree after reaching either 25, 75 or 150 taxa.
For each tree, we arbitrarily set the outgroup to be the first taxa (alphabetically) thus effectively rooting the trees on this taxa.
These trees are hereafter called the ``true'' trees.
We then simulated discrete morphological characters on the topology of these trees using the either of the two following models:
\begin{itemize}
    \item The morphological HKY-binary model \citep{OReilly20160081} which is an HKY model \citep{HKY85} with a random states frequency (sampled from a uniform distribution $(0,1)$ and scaled to sum to $1$) and using a transition/transvertion rate of $2$ \citep{douadycomparison2003} but where the purines (A,G) were changed into state $0$ and the pyrimidines (C,T) in state $1$.
    This model has the advantage of not favouring Bayesian inference \citep[since it doesn't use a M$k$ model;][; see discussion]{OReilly20160081} but the downside of it is it can only generate binary state characters \citep[or 4 states;][]{puttick2017uncertain}.
    \item To generate more than binary states characters, we used the M$k$ model \citep{lewisa2001}.
    We draw the number of character states with a probability of $0.85$ for binary characters and $0.15$ for three state characters \citep{Guillerme2016146,ZouConvergence}.
    This model assumes a equal transition rate between character states that might seem overly simplifying and excludes other observed transition patterns \citep[e.g. Dollo characters;][]{Dollo,wright2015came}.
    Recently however, \cite{Wright01072016} have shown that an equal rate transition is still the most present in empirical data.
\end{itemize}

@@

This being said, it is interesting to analysis this difference in tree inference method in the light of the effect of high/low character differences.
In fact the differences observed here could be due to the inherent mechanisms of each method.
When many characters are homoplasic (high character difference) parsimony will cope with this information by generating longer trees (which will be more likely to be discarded in the tree search).
Conversely, Bayesian analysis can (1) cope with homoplasy more easily by increasing branch length to accommodate more changes and (2) will also incorporate sub-optimal trees in the final results.
These two points could explain the fact that the effect of character correlation seems less important in Bayesian inference, especially in terms of clade conservation.

>>>>

The last two decades have witnessed a resurgence of interest in the use of morphological character data in phylogenetic studies.
This owes in large part to the use of fossils to undertake at least partial reconstructions of phylogenetic trees, especially where ancestral states reconstructions or absolute calibrations of divergence times are necessary. 
Morphological character data are often considered inferior to molecular sequence data (due to their inherent biases), but are often the only source of data---especially for fossils.
While there is a general appreciation of the limits of morphological data, they are frequently dismissed without any empirical investigations into their statistical properties.
As morphological data are likely to continue to play an extensive role in phylogenetic analysis, it is essential to understand the circumstances under which morphological data might be expected to 'misbehave'.
This opens up possibilities for predicting problematic datasets and possibly proposing new confidence measures in phylogenetic datasets.
%The last decade has witnessed a resurgence of use of discrete morphological data in macroevolutionary studies.
%Traditionally confined to ``classic'' cladistic studies in palaeontology, these methods have recently been widely expanded and improved whether based on the models and methods for inferring phylogenies \citep[e.g.][]{heath2014fossilized,Wright01072016} or to combine them with molecular data \citep[e.g.][]{pyrondivergence2011,ronquista2012}.
%In parallel, much development has also been done on the pre-phylogenetic analysis \citep[e.g. data collection;][]{morphobank} as well as on post-phylogenetic ones \citep[e.g. morphological disparity analysis;][]{Close2015,Claddis}.
%Nonetheless, using discrete morphological data in macroevolutionary studies comes with several caveats based to the nature of data used \citep{Guillerme2016146,bapst2017combined}; the coding of this data \citep{Brazeau2011,simoes2017giant}; the topology inference method \citep{spencerefficacy2013,wrightbayesian2014,OReilly20160081,puttick2017uncertain,goloboff2017weighted} or the dating methods \citep{Arcila2015131,o2016tips}.

@@

    \item \textbf{Evolutionary dependence:} this is the result of sets of characters co-evolving: the character's correlation is thus due to their shared evolutionary history.
% MDB: Does the above really describe this correctly. This actually sounds a lot more like genuine synapomorphy than potentially problematic correlation.
%TG: Yep, totally synapormophies. Although there is a subtle nuance in the way they are studied. In pure phylo approach (i.e. matrix -> tree), the synapomorphic characters are participating to the topology inference and are thus impossible to distinguish (i.e. the synapomorphies make the trees and the trees make the synapomorphies). In the papers cited just below they have a different approach. They assume a phylogeny and then "map" characters on to it. In this later case, the synapomorphies are effectively decoupled from the topology.
    Many methods have been developed to study these correlations, especially since they can provide us with a lot of information on how specific groups acquired specific characteristics \citep{Lande1983,Maddison1990,Pagel1994,Pagel2006,Grabowski2016}.
    % For example, the apparition of the jaw as a character in gnatosthomes correlates with a whole suite of characters present in modern gnatosthomes (e.g. beaks, tribosphenic molars, jaw protrusion, etc.). %TG: this links to inapplicable data!
%MDB: The above is really coding dependence that you describe below, in a way. It's a character hierarchy. What we would like to discuss here is probably something more like limb reduction and axial elongation, for instance. These two things seem to co-vary extensively in tetrapods, with snake-like body forms evolving multiple times in numerous tetrapod lineages.
%TG: good call, I've changed the example.
    For example, in vertebrates, axial elongation can be correlated to limb reduction with snake-like bodies evolving multiple times in numerous tetrapod lineages.

@@

\noindent Where $n$ is the number of taxa with comparable data and $x_i$ and $y_i$ are each characters states for the characters $x$ and $y$ and the taxa $i$.
$CD$ is a continuous distance metric bounded between $0$ and $1$ (see the mathematical demonstration in the Supplementary Material 1).
Since we are considering differences as being only Fitch-like (i.e. non-weighted and non-ordinate), we calculated the difference between character states in a qualitative way.
Two same character states tokens have a difference of zero and two different ones have a difference of one (e.g. $0 - 0 = 0$ or $1 - 8 = 1$).
Additionally, we only consider differences for taxa with shared information \citep[i.e. a Gower distance;][]{GowerDist}.
We standardised each character by arbitrarily modifying their character state tokens by order of appearance.
In other words, we replaced all the occurrences of the first token to be $1$, the second to be $2$, etc.
This way, a character \texttt{A = \{2,2,3,0,0,3\}} for six taxa would be standardised as \texttt{A' = \{1,1,2,3,3,2\}} \citep[following the \textit{xyz} notation in][p.13]{felsenstein2004inferring}.
Note that in terms of phylogenetic signal, both \texttt{A} and \texttt{A'} are exactly identical (forming three distinct groups).

@@

\subsection{Simulating discrete morphological matrices}
To simulate the matrices we applied a protocol very similar to \cite{Guillerme2016146}.
First, we generate random birth-death trees with the birth ($\lambda$) and death ($\mu$) parameters sampled from a uniform $(0,1)$ distribution maintaining $\lambda$ $>$ $\mu$ using the \texttt{diversitree} \texttt{R} package \citep[v0.9-8;][]{fitzjohndiversitree2012} and saving the tree after reaching either 25, 75 or 150 taxa.
For each tree, we arbitrarily set the outgroup to be the first taxa (alphabetically) thus effectively rooting the trees on this taxa.
These trees are hereafter called the ``true'' trees.
We then simulated discrete morphological characters on the topology of these trees using the either of the two following models:
\begin{itemize}
    \item The morphological HKY-binary model \citep{OReilly20160081} which is an HKY model \citep{HKY85} with a random states frequency (sampled from a uniform distribution $(0,1)$ and scaled to sum to $1$)
    %TG: is that not a random Dirichlet? D(A,G,C,T)?
     and using a transition/transvertion rate of $2$ \citep{douadycomparison2003} but where the purines (A,G) were changed into state $0$ and the pyrimidines (C,T) in state $1$.
    This model has the advantage of not favouring Bayesian inference \citep[since it doesn't use a M$k$ model;][; see discussion]{OReilly20160081} but the downside of it is it can only generate binary state characters \citep[or 4 states;][]{puttick2017uncertain}.
    \item To generate more than binary states characters, we used the M$k$ model \citep{lewisa2001}.
    We draw the number of character states with a probability of $0.85$ for binary characters and $0.15$ for three state characters \citep{Guillerme2016146,ZouConvergence}.
    This model assumes a equal transition rate between character states that might seem overly simplifying and excludes other observed transition patterns \citep[e.g. Dollo characters;][]{Dollo,wright2015came}.
    Recently however, \cite{Wright01072016} have shown that an equal rate transition is still the most present in empirical data.
\end{itemize}

@@

This being said, it is interesting to analyse this difference in tree inference method in the light of the effect of high/low character differences.
In fact the differences observed here could be due to the inherent mechanisms of each method.
When many characters are homoplasic (high character difference) parsimony will cope with this information by generating longer trees (which will be more likely to be discarded in the tree search).
Conversely, Bayesian analysis can (1) cope with homoplasy more easily by increasing branch length to accommodate more changes and (2) will also incorporate sub-optimal trees in the final results.
These two points could explain the fact that the effect of character correlation seems less important in Bayesian inference, especially in terms of clade conservation.
